# Descriptive Analytics

Copy the BitCoin data frame to a new data frame named BitCoin_df. In the new data frame create two more columns ‘month’ & ‘year’ by populating with the months & years values from the ‘Date’ column. 
```{r, echo=FALSE}


#("tidyverse")
#Load the dataset from csv into a data frame named BitCoin.
library(tidyverse)

#load csv from github(uploaded for Assignment_02_Group_04)
BitCoin <- read.csv("https://raw.githubusercontent.com/shakibed/Final_Project_Group_04/main/BTC-Monthly.csv", stringsAsFactors = FALSE)

#convert to date
BitCoin$Date <- as.Date(BitCoin$Date)

#it shows no missing value found
# Copy the bitcoin data frame to a new data frame named BitCoin_df
BitCoin_df<- BitCoin


# Create 'month' and 'year' columns in the BitCoin_df data frame
BitCoin_df <- BitCoin_df %>%
  mutate(
    month = format(Date, "%m"),
    year = format(Date, "%Y")
  )

```
 
  - Create a monthly boxplot of prices.
  - Create a yearly boxplot of prices.
  - Create year wise trend lines of prices.
  -- Convert the BitCoin data frame to a time series object with frequency 1.
  - Plot the time series of monthly prices on years. Use blue color dots to plot the data points. Connect the points with a line.
  - Find the relationship between consecutive days. Show the correlation through a scatter plot.
  
 
## Create a monthly boxplot of prices.{-}


```{r}
# 1. Create a monthly boxplot of prices

ggplot(BitCoin_df, aes(x = factor(month), y = Close, fill = factor(month))) +
  geom_boxplot(outlier.colour = "red") +
  labs(title = "Monthly Boxplot of Bitcoin Prices", x = "Month", y = "Close Price") +
  scale_fill_viridis_d(name = "Month", labels = month.abb)


```

_**Code explanation here**_

* The boxplot shows that Bitcoin prices exhibit some level of seasonal variation. 
* Prices tend to be higher during certain months (March, April, May) compared to others ( June, July). 
* This could indicate seasonal trends or cycles in the market.
* The presence of outliers (red dots) in some months, especially in May and September, indicates that there were extreme price changes during these periods.


## Create a yearly boxplot of prices.{-}


```{r}
library(RColorBrewer)

# 2. Create a yearly boxplot of prices
color_palette <- brewer.pal(9, "Set3")  # 9 colors for 9 years

#the boxplot with a legend for the years
ggplot(BitCoin_df, aes(x = factor(year), y = Close, fill = factor(year))) +
  geom_boxplot(outlier.colour = "red") +
  labs(title = "Yearly Boxplot of Bitcoin Prices", x = "Year", y = "Close Price") +
  scale_fill_manual(name = "Year", values = setNames(color_palette, as.character(2015:2023))) +
  theme(legend.position = "right")


```


_**Code explanation here**_


## Create year wise trend lines of prices.{-}

* A boxplot showing the distribution of Bitcoin closing prices over different years. 
* Each box represents the interquartile range (IQR) of the closing prices for a particular year, with the horizontal line inside the box indicating the median price. T
* he whiskers extend to the minimum and maximum values within 1.5 times the IQR from the first and third quartiles, respectively. 
* Outliers beyond this range are shown as individual points.



```{r}
# 3. Create year-wise trend lines of prices
ggplot(BitCoin_df, aes(x = Date, y = Close, color = year)) +
  geom_line() +
  labs(title = "Year-Wise Trend Lines of Bitcoin Prices", x = "Date", y = "Close Price") +
  theme(legend.position = "bottom")


```

_**Explanation of Plot**_

* A trend line showing the price of Bitcoin changes over different years. 
* 2015,2016: Bitcoin prices were relatively low and stable, with prices mostly below $1,000.
* 2017: Significant growth was observed, with prices starting below $1,000 and rising to nearly $20,000 by the end of the year.
* 2018,2019: After the peak in 2017, prices dropped significantly, fluctuating between $3,000 and $10,000.Prices started to recover, stabilizing between $3,000 and $12,000.
* 2020: There was a steady increase, with prices rising from around $5,000 to over $30,000 by the end of the year.
* 2021: Marked a dramatic rise, with Bitcoin prices reaching new all-time highs, peaking around $60,000.
* 2022: Prices were highly volatile, with significant drops and recoveries, ranging from $30,000 to $50,000.
* 2023: Prices appeared to stabilize somewhat, fluctuating between $20,000 and $40,000.



### Convert the BitCoin data frame to a time series object with frequency 1{-}
```{r}

# Convert the BitCoin data frame to a time series object with frequency 1
bitcoin_ts <- ts(BitCoin_df$Close, start = c(as.numeric(format(min(BitCoin_df$Date), "%Y")), 1), frequency = 1)

# Print the time series object
print(bitcoin_ts)
```


### Plot the time series of monthly prices on years. Use blue color dots to plot the data points. Connect the points with a line.{-}


```{r}
# Plot the time series of monthly prices on years
ggplot(BitCoin_df, aes(x = Date, y = Close)) +
  geom_point(color = "blue") + # Blue color dots for data points
  geom_line(color = "blue") +  # Connect the points with a line
  labs(title = "Time Series of Monthly Bitcoin Prices",
       x = "Year",
       y = "Close Price") +
  theme_minimal()


```

_**Explanation of Plot**_

Explanation of the Plot:
- Horizontal Axis (Year): Represents the timeline from 2016 to 2024.
- Vertical Axis (Close Price): Represents the closing price of Bitcoin at the end of each month.
- Blue Dots: Each dot represents the closing price of Bitcoin for a particular month.
- Blue Line: Connects the dots to show the trend and continuity of Bitcoin prices over time.
- The plot provides a visual representation of Bitcoin's price volatility and trends, making it easier to analyze patterns and potential factors affecting its value over the given period.

_**Result**_

The plot reveals significant fluctuations in Bitcoin prices, including notable peaks around 2021 and subsequent declines. The most recent data indicates a rising trend in prices again.


## Find the relationship between consecutive months. Show the correlation through a scatter plot.{-}



```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

# Aggregate the data by month
BitCoin_df_monthly <- BitCoin_df %>%
  mutate(YearMonth = floor_date(Date, "month")) %>%
  group_by(YearMonth) %>%
  summarise(Monthly_Close = last(Close)) %>%
  arrange(YearMonth)

# Create a new column for the previous month's closing price
BitCoin_df_monthly <- BitCoin_df_monthly %>%
  mutate(Prev_Month_Close = lag(Monthly_Close, 1))

# Remove the first row with NA value in Prev_Month_Close
BitCoin_df_monthly <- BitCoin_df_monthly %>%
  filter(!is.na(Prev_Month_Close))

# Calculate the correlation
monthly_correlation <- cor(BitCoin_df_monthly$Monthly_Close, BitCoin_df_monthly$Prev_Month_Close)
print(paste("Correlation between consecutive months: ", monthly_correlation))

# Create a scatter plot
ggplot(BitCoin_df_monthly, aes(x = Prev_Month_Close, y = Monthly_Close)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of Consecutive Month Prices",
       subtitle = paste("Correlation: ", round(monthly_correlation, 2)),
       x = "Previous Month Close Price",
       y = "Current Month Close Price") +
  theme_minimal()
```

_**Code explanation here**_

The scatter plot provides a clear visual representation of the relationship between Bitcoin prices in consecutive months, highlighting the strong correlation 0.96 and linear trend, which are valuable for analyses and decision-making processes.

- Horizontal Axis (Previous Month Close Price): Represents the closing price of Bitcoin for the previous month.
- Vertical Axis (Current Month Close Price): Represents the closing price of Bitcoin for the current month.
- Blue Dots: Each dot represents the closing price of Bitcoin for a specific month plotted against the closing price of the previous month.
- Correlation Value: Displayed at the top of the chart, it quantifies the strength of the relationship between - consecutive month's prices.

