--- 
title: "Final Project-Group 04"
author: "Shakib-241001661,Shudeshna-241000261,Mithun-241002261,Pradipto-241002161,Mufrad-241001461,Shahriar Islam-241000461"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "Regression & Time Series analysis and forecasting techniques on Bitcoin historical price."
---

# Objective of the Project{-}

perform **Regression** & **Time Series analysis** and **forecasting techniques** on _Bitcoin_ historical price. The experience can be replicated to forecast Stock price & Currency price as well. 

 
```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```
 
## Dataset{-}

A dataset of **Bitcoin** (BTC) historical price against USD. This is a monthly average price dataset having prices from 01 January 2015 to 30 November 2023. 

## Applied Model{-}
  1.	Linear regression model
  2.	Quadratic regression model
  3.	ARMIA model
 
 

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
